Getting Started with Multilevel Modeling in R
========================================================

## Jared E. Knowles

# Introduction

Analysts dealing with grouped data and complex hierarchical structures in their 
data ranging from measurements nested within participants, to counties nested within 
states or students nested within classrooms often find themselves in need of modeling 
tools to reflect this structure of their data. In R there are two predominant ways 
to fit multilevel models that account for such structure in the data. These tutorials 
will show the user how to use both the `lme4` package in R to fit linear and nonlinear 
mixed effect models, and to use `rstan` to fit fully Bayesian multilevel models. The 
focus here will be on how to fit the models in R and not the theory behind the models. 
For background on multilevel modeling, see the references. [1]

This tutorial will cover getting set up and running a few basic models using `lme4` 
in R.Future tutorials will cover:

* constructing varying intercept, varying slope, and varying slope and intercept models in R
* generating predictions and interpreting parameters from mixed-effect models
* generalized and non-linear multilevel models
* fully Bayesian multilevel models fit with `rstan` or other MCMC methods

# Setting up your enviRonment

Getting started with multilevel modeling in R is simple. `lme4` is the canonical 
package for implementing multilevel models in R, though there are a number of packages 
that depend on and enhance its feature set, including Bayesian extensions. `lme4` 
has been recently rewritten to improve speed and to incorporate a C++ codebase, and 
as such the features of the package are somewhat in flux. Be sure to update the package 
frequently. 

To install `lme4`, we just run:

```{r eval=FALSE, results='hide', echo=TRUE}
# Main version
install.packages("lme4")

# Or to install the dev version
library(devtools)
install_github("lme4",user="lme4") 
```

```{r setup, echo=FALSE, error=FALSE, message=FALSE, eval=TRUE, results='hide'}
opts_chunk$set(dev='svg', fig.width=6, fig.height=6, echo=TRUE, 
               message=FALSE, error=FALSE)

```

# Read in the data

Multilevel models are appropriate for a particular kind of data structure where 
units are nested within groups (generally 5+ groups) and where we want to model 
the group structure of the data. For our introductory example we will start with 
a simple example from the `lme4` documentation and explain what the model is doing. 

```{r loadandviewdata}
data(sleepstudy)

str(sleepstudy)

summary(sleepstudy)

```

Here we have data on reaction times of subjects in a study about sleep. Each subject 
is observed ten times - once per day. Thus we have observations nested within subjects. 

# Fit the Non-Multilevel Models



[1] Examples include Gelman and Hill, Gelman et al. 2013, etc. 